A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-12-19 podc_paper/garbage-collection.tex
init garbage_collection, add pdf

\section{Feilds}
\begin{algorithm}
\caption{Tree Fields Description}
\begin{algorithmic}[1]
\setcounter{ALG@line}{1}

@@

\section{Queue}
\begin{algorithm}
\caption{\tt{\sl{Queue}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{0}

@@

\Statex $\triangleright$ Returns the response to $D_i(root,b)$, the \nf{i}th \nf{Dequeue} in \nf{root.blocks[b]}.
\Function{element}{FindResponse}{\sl{int} b, \sl{int} i}
\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
\State \tt{ \color{red} lastDequeudFrom.update(b)}
\State \Return \tt{null} \label{returnNull}
\Else \cmt{The response is $E_e(root)$, the \nf{e}th \nf{Enqueue} in the root.}
\State \tt{e= i + (root.blocks[b-1].sum\sub{enq}-root.blocks[b-1].size)} \label{computeE}
\State {\color{red}\tt{<x, y>= root.\Call{DoublingSearch}{e, b}}}
\State {\color{red}\tt{lastDequeudFrom.update(x)}}
\State \Return \tt{root.GetEnqueue(x,y)}\label{findAnswer}
\EndIf
\EndFunction{FindResponse}

@@

\section{Search+Append}
\begin{algorithm}
\caption{\tt{\sl{Node}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{25}

@@

\begin{algorithm}
\caption{\tt{\sl{Root}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{36}
\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})=E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within $\triangleright$~the \nf{b}th block in the \nf{root}.}

\Function{<int, int>}{DoublingSearch}{\sl{int} e, \sl{int} end} \cmt{{\color{red} I think with garbage collection, doubling search is not needed any more and a binary search on \nf{root.blocks} would be good enough.}}
% \State \tt{start= end-1} \label{dsearchStart} 
% \While{\tt{root.blocks[start].sum\sub{enq}}$>=$\tt{e}}
% \State \tt{start= max(start-(end-start), 0)} \label{doubling}
% \EndWhile \label{dsearchEnd}
% \State \tt{b= root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
% \State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
% \State\Return \tt{<b,i>}
\EndFunction{DoublingSearch}
\end{algorithmic}
\end{algorithm}

@@

\section{Propagate}
\begin{algorithm}
\caption{\tt{\sl{Node}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{50}

@@

\Statex $\triangleright$ \textmd{Creates a block containing new operations of \nf{this.}children, and then tries to append it to \nf{this}.}
\Function{boolean}{Refresh()}{}
\State \tt{h= \head} \label{readHead}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
\If{\nf{dir.blocks[h\sub{dir}]!=null}} \label{ifHeadnotNull}
\State{\tt{dir.\Call{Advance}{h\sub{dir}}}} \label{helpAdvance}
\EndIf
\EndFor \label{endHelpChild1}
\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
\EndIf
\State{\tt{\color{red} result= blocks.TryAppend(new, h)}} \label{cas}
\State{\tt{this.\Call{Advance}{h}}} \label{advance}
\State \Return{ \tt{result}}

@@

\section{Index+Get}

@@

\color{red}
\section{MaxByProcess}

@@

\section{Help}

@@

\section{FreeMemory}

@@

\section{PBRT}

@@

\Statex
\Function{}{TryAppend}{\sl{block} \tt{B}, \sl{int} \tt{i}}
\Statex {\color{red} Tries to append \nf{B} with key \nf{i} to the red-black tree.}
{\color{red}\If{\tt{i\%p\textsuperscript{2}=0}}
\State \Call{Help()}{}
\State \tt{root.}\Call{FreeMemory()}{lastDequeuedFrom.Get()-1}
\State \tt{garbageCollectRound=floor(root.head/p\textsuperscript{2})}
\EndIf}
\EndFunction{FreeMemory}

@@

\pagebreak
\section{Description}

\color{black}
In our original algorithm an \nf{Enqueue} or a \nf{Dequeue} remains in the \nf{blocks} array in the tree nodes even after they terminate. This makes the space used by the algorithm factor of the number of operations of invoked on the queue. In this section here we want to free the memory allocated by the operations that are no longer needed and make the space used polynomial of $p+q$.

One way of handling garbage collection is to reallocate the space taken by each operation right after it is not needed anymore, but it might cost too much to do that. So we garbage collect by batching. If we garbage collect the unnecessary blocks in a node every $p^2$ block appended to the node, the garbage collection cost is amortized over $p^2$ blocks which is $O(p^3)$ and $\Omega(p^2)$.

In our design a process garbage collects a node when it attempts to append a block in the $kp^2$ position in the. \nf{Garbage Collect} corresponded to the $kp^2$th block in node $n$ is called the $k$th round of garbage collect on $n$. Every $p^2$  block appended to a node at least one garbage collect happens because the node's \nf{head} cannot advance until the garbage collect is done (see Lines ).

\begin{lemma}
The number of the blocks in the \nf{blocks} is $O(p^2+q)$.
\end{lemma}

An \nf{Enqueue} operation can be removed from the tree after its \nf{element} has been computed to be the response of some \nf{Dequeue}. It is safe to remove a \nf{Dequeue} operation from the tree after the \nf{Dequeue} is terminated, but a \nf{Dequeue} might sleep for a long time and prevents to be garbage collected. In that situation other processes can help the \nf{Dequeue} to compute its response and wtite down its response somewhere. If we remove a \nf{Dequeue} from the tree after it is helped the \nf{Dequeue} can read the helped response writtern when it encountered a problem (computing the \nf{Dequeue} index in the \nf{root} or getting the response \nf{Enqueue}). A \nf{Dequeue} operation can be removed from the tree after its response has been computed. We say a block  is \it{finished} if all of its operations can be removed.

\begin{lemma}
There are at most $O(p^2+q)$ unfinished blocks in a node's \nf{blocks} at a time.
\end{lemma}

If the $i$th \nf{Enqueue} gets dequeued in a FIFO queue, it means $1$ to $i-1$th \nf{Enqueue} operations are also dequeued. This gives us the idea that if a block is finished then all the blocks before it are also finished. If an operation in a block goes to sleep for a long time then other processes help the operation so the block get finished. Note that when an \nf{Enqueue} operation in a block is not finished the block cannot be finished until some \nf{Dequeue} dequeues that \nf{Enqueue}. Since there are at most $p$ idle operations, we can help them before garbage collection and then remove all the finished blocks safely.

@@

>>>>

\begin{algorithm}
\caption{Tree Fields Description}
\begin{algorithmic}[1]
\setcounter{ALG@line}{1}

@@

\begin{algorithm}
\caption{\tt{\sl{Queue}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{0}

@@

\Statex $\triangleright$ Returns the response to $D_i(root,b)$, the \nf{i}th \nf{Dequeue} in \nf{root.blocks[b]}.
\Function{element}{FindResponse}{\sl{int} b, \sl{int} i}
\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
\State \tt{ \color{red} lastDequeudFrom.update(b)}
\State \Return \tt{null} \label{returnNull}
\Else \cmt{The response is $E_e(root)$, the \nf{e}th \nf{Enqueue} in the root.}
\State \tt{e= i + (root.blocks[b-1].sum\sub{enq}-root.blocks[b-1].size)} \label{computeE}
\State {\color{red}\tt{<x, y>= root.\Call{BinarySearch}{e}}}
\State {\color{red}\tt{lastDequeudFrom.update(x)}}
\State \Return \tt{root.GetEnqueue(x,y)}\label{findAnswer}
\EndIf
\EndFunction{FindResponse}

@@

\begin{algorithm}
\caption{\tt{\sl{Node}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{25}

@@

@@

\begin{algorithm}
\caption{\tt{\sl{Node}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{50}

@@

\Statex $\triangleright$ \textmd{Creates a block containing new operations of \nf{this.}children, and then tries to append it to \nf{this}.}
\Function{boolean}{Refresh()}{}
\State \tt{h= \head} \label{readHead}
{\color{red}\If{\tt{h\%p\textsuperscript{2}=0}}
\State \Call{Help()}{}
\State \tt{root.}\Call{FreeMemory()}{lastDequeuedFrom.Get()-1}
\EndIf}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
\If{\nf{dir.blocks[h\sub{dir}]!=null}} \label{ifHeadnotNull}
\State{\tt{dir.\Call{Advance}{h\sub{dir}}}} \label{helpAdvance}
\EndIf
\EndFor \label{endHelpChild1}
\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
\EndIf
\State{\tt{\color{red} result= blocks.TryAppend(new, h)}} \label{cas}
\State{\tt{this.\Call{Advance}{h}}} \label{advance}
\State \Return{ \tt{result}}

@@

@@

\color{red}

@@

@@

@@

@@

@@

\pagebreak
\clearpage
\section{Description}

\color{black}
In our algorithm an \nf{Enqueue} or a \nf{Dequeue} remains in the \nf{blocks} array in the tree nodes even after they terminate. This makes the space used by the algorithm factor of the number of operations of invoked on the queue. Here in this section, we want to add a mechanism to free the memory allocated by the operations that are no longer needed and make the overall space used polynomial of $p+q$.

It is a common way to do garbage collection in batches. If we garbage collect the unnecessary blocks in the nodes every $p^2$ block appended to the root, the garbage collection cost is amortized over $p^2$ blocks which is $O(p^3)$ and $\Omega(p^2)$ operations.

In our design a process attempts to collect the garbage, when it is going to to append a block in the $kp^2$ position in \nf{root.blocks} (see Line \ref{}). Every $p^2$  block appended to the root, one \nf{GarbageCollect} terminates because \nf{root.head} cannot advance until a \nf{GarbageCollect} garbage collect is done (see Lines ).

\begin{lemma}
Size of \nf{root.blocks} after \nf{GarbageCollect} is $O(p^2+q)$.
\end{lemma}

\begin{lemma}
Total number of the blocks in the tree is $O(p^3+pq)$.
\end{lemma}

\nf{Enqueue} operation $e$ can be removed from the tree after termination of \nf{Dequeue} $d$ where $Resp(d)=e$ or $e.$\nf{element} has been computed to be $RESP(d)$. It is safe to remove \nf{Dequeue} $d$o from the tree after the $d$ is terminated or $Resp(d)$\nf{.element} has been computed. We can remove a block after the told conditions are satisfied for all ofits operations. A \nf{Dequeue} may go to sleep for a long time and prevent the \nf{block} to be removed. In that situation other processes can help the \nf{Dequeue} by computing its response and writing it down somewhere. After writing down the response of th \nf{Dequeue} it is safe to remove. Because if the \nf{Dequeue} failed to compute its response it can read the response written (computing the \nf{Dequeue} index in the \nf{root} or getting the response \nf{Enqueue}).

\begin{definition}
A block  is \it{finished} if all of its \nf{Enqueue}s have been computed to be the response of some \nf{Dequeue} and all of its \nf{Dequeues} responses have been computed.  
\end{definition}

\begin{corollary}
If a block is finished, then all of its subblocks are also finished.  
\end{corollary}

\begin{lemma}
It is safe to remove a finished block from the tree.
\end{lemma}

If the $i$th \nf{Enqueue} gets dequeued in a FIFO queue, it means the first $i-1$ \nf{Enqueue} operations have been already dequeued. This gives us the idea that if a block is finished then all the blocks before it are also finished. If an operation in a block goes to sleep for a long time then other processes help the operation so the block get finished. Note that when an \nf{Enqueue} operation in a block is not finished the block cannot be finished until some \nf{Dequeue} dequeues that \nf{Enqueue}. Since there are at most $p$ idle operations, we can help them before garbage collection and then remove all the finished blocks safely.

@@

\begin{lemma}
  The amortized step per process for \nf{GarbageCollect} is $O()$.
\end{lemma}

\begin{lemma}
  Algorithm is wait-free and linearizable.
\end{lemma}
